---
title: "Variety-Trial Model Validation"
author: "AJ"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: readable
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = FALSE,
  message    = FALSE,
  warning    = FALSE,
  fig.align  = "center",
  fig.width  = 8,
  fig.height = 5
)

suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(readr)
  library(tidyr)
  library(purrr)
  library(stringr)
  library(gridExtra)
  library(tibble)
})
```

# Purpose

This document validates that the new spatial + multi-environment (MET) workflow estimates what we intend and quantifies how well it performs, using the simulated variety-trial data.

**We check:**

1. Recovery of entry effects and ranks versus known simulation truth.  
2. Stage‑1 spatial model selection behavior.  
3. Uncertainty calibration and practical interpretability (CIs, Tukey letters).  
4. Optional diagnostics on variance components and residual structure.

---

# Data inputs and file structure

This validation assumes the Rmd file lives in the project’s `code/` folder and the backend has been run with `DATA_MODE = "sim"` so that all simulated outputs are available.

Expected files (relative to the project root):

- `sim_data/trial_sim.csv`: simulated plots with columns `site, year, env, rep, row, col, entry, yield`.  
- `sim_data/entry_truth.csv`: known entry main effects used to simulate data.  
- `sim_output/lsm_stage1.csv`: per‑environment LS‑means and SEs after spatial + fallback modeling.  
- `sim_output/lsm_across.csv`: across‑environment LS‑means (MET stage‑2).  
- `sim_output/diffs_across.csv`: Tukey‑adjusted pairwise comparisons across environments.  
- `sim_output/cld_across.csv`: compact letter display (“grouping letters”) across environments.  
- `sim_output/entry_blups.csv`: BLUPs of entry effects from the MET model.  
- `sim_output/run_report.txt`: textual summary of modeling choices and assumptions.

```{r load-data}
# Paths are relative to the project *root*, assuming this Rmd is in ./code
trial_path   <- "../sim_data/trial_sim.csv"
truth_path   <- "../sim_data/entry_truth.csv"
s1_path      <- "../sim_output/lsm_stage1.csv"
lsm_ac_path  <- "../sim_output/lsm_across.csv"
diffs_path   <- "../sim_output/diffs_across.csv"
cld_path     <- "../sim_output/cld_across.csv"
blups_path   <- "../sim_output/entry_blups.csv"
report_path  <- "../sim_output/run_report.txt"

need <- c(trial_path, truth_path,
          s1_path, lsm_ac_path,
          diffs_path, cld_path,
          blups_path)

missing <- need[!file.exists(need)]
if (length(missing) > 0) {
  stop(
    paste(
      "Missing files:",
      paste(missing, collapse = "\n  "),
      "\nRun the simulator and analyze_trial() with DATA_MODE = 'sim' before knitting this validation report."
    )
  )
}

trial   <- readr::read_csv(trial_path,  show_col_types = FALSE)
truth   <- readr::read_csv(truth_path,  show_col_types = FALSE)
s1      <- readr::read_csv(s1_path,     show_col_types = FALSE)
lsm_ac  <- readr::read_csv(lsm_ac_path, show_col_types = FALSE)
diffs   <- readr::read_csv(diffs_path,  show_col_types = FALSE)
cld     <- readr::read_csv(cld_path,    show_col_types = FALSE)
blups   <- readr::read_csv(blups_path,  show_col_types = FALSE)

# Normalize entry labels to character everywhere
truth  <- truth %>% mutate(entry = as.character(entry))
lsm_ac <- lsm_ac %>% mutate(entry = as.character(entry))
blups  <- blups  %>% mutate(entry = as.character(entry))
s1     <- s1     %>% mutate(entry = as.character(entry))

# Optional: load run_report if present
run_report <- if (file.exists(report_path)) {
  readLines(report_path, warn = FALSE)
} else {
  NULL
}
```

---

# 1. Recovery of entry effects and ranks

**Goal.** Across environments, entries’ adjusted means and BLUPs should track the true entry effects used in simulation.

**Metrics reported.** Spearman rank correlation (ranking fidelity) and RMSE (numerical accuracy).

```{r recovery-metrics}
cmp_lsm  <- truth %>% inner_join(lsm_ac,  by = "entry")
cmp_blup <- truth %>% inner_join(blups,   by = "entry")

metr <- tibble(
  source   = c("Across-env LS-means", "BLUPs"),
  rank_cor = c(
    cor(cmp_lsm$effect, cmp_lsm$estimate, method = "spearman"),
    cor(cmp_blup$effect, cmp_blup$BLUP,   method = "spearman")
  ),
  rmse     = c(
    sqrt(mean((cmp_lsm$estimate - cmp_lsm$effect)^2)),
    sqrt(mean((cmp_blup$BLUP     - cmp_blup$effect)^2))
  )
)

metr
```

Higher rank correlation indicates better retention of ordering among entries, while lower RMSE indicates closer numerical recovery. In many MET settings, BLUPs stabilize ranks under unbalanced or noisy data, so it is informative to compare both summaries.

```{r scatter-plots, fig.height=4.5}
p1 <- ggplot(cmp_lsm, aes(x = effect, y = estimate)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Truth vs across-environment LS-means",
    x = "True entry effect",
    y = "LS-mean estimate"
  ) +
  theme_bw()

p2 <- ggplot(cmp_blup, aes(x = effect, y = BLUP)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Truth vs entry BLUPs",
    x = "True entry effect",
    y = "BLUP"
  ) +
  theme_bw()

gridExtra::grid.arrange(p1, p2, nrow = 1)
```

---

# 2. Stage‑1 spatial model selection

**Goal.** Confirm that, per environment, AICc selects plausible covariance families and that fallbacks are triggered only when needed.

```{r stage1-cov-summary}
cov_tab <- s1 %>%
  count(env, cov) %>%
  arrange(env, desc(n))

cov_freq <- s1 %>%
  count(cov) %>%
  mutate(prop = n / sum(n))

list(
  "Chosen covariance per environment" = cov_tab,
  "Overall covariance frequency"      = cov_freq
)
```

Under anisotropic simulation, `expa` is often selected. If the simulator uses isotropic structure, `exp`, `sph`, or `gau` may be competitive. The table also reveals when environments fell back to `none_rcbd` or `none_lm` due to failed spatial fits or limited replication.

---

# 3. Uncertainty calibration

**Goal.** Per‑environment LS‑mean 95% confidence intervals should cover the true effect at approximately 95%.

```{r coverage}
# Center each environment’s LS-means and compare to entry effects
s1_centered <- s1 %>%
  group_by(env) %>%
  mutate(estimate_c = estimate - mean(estimate, na.rm = TRUE)) %>%
  ungroup()

cov_df <- s1_centered %>%
  select(env, entry, estimate_c, stderr) %>%
  inner_join(truth, by = "entry") %>%
  mutate(
    lo       = estimate_c - 1.96 * stderr,
    hi       = estimate_c + 1.96 * stderr,
    covered  = (effect >= lo) & (effect <= hi)
  )

mean(cov_df$covered, na.rm = TRUE)
```

Coverage near 0.95 suggests that interval estimates are reasonably calibrated. Substantially lower coverage may indicate residual variance underestimation or a misspecified spatial structure.

As an additional rank‑based check, we can compare truth to the across‑env LS‑means after aligning their overall location.

```{r coverage-rank-check}
offset <- mean(lsm_ac$estimate, na.rm = TRUE) -
  mean(truth$effect, na.rm = TRUE)

with(dplyr::inner_join(truth, lsm_ac, by = "entry"),
     cor(effect + offset, estimate))
```

---

# 4. Practical separation: Tukey letters and pairwise differences

**Goal.** Summarize Tukey’s multiple comparisons using compact letter displays (CLD). Entries sharing a letter are **not** significantly different at the chosen alpha.

```{r cld-merge, fig.height=5}
# Ensure CLD has a 'group' column (can arrive as .group)
if ("group" %in% names(cld)) {
  cld_fix <- cld
} else if (".group" %in% names(cld)) {
  cld_fix <- dplyr::rename(cld, group = .group)
} else {
  stop("CLD table must contain a 'group' or '.group' column.")
}

# Standardize CI column names in case they have trailing dots
fix_ci_names <- function(df){
  nm <- names(df)
  nm <- sub("^lower\\.CL\\.$", "lower.CL", nm)
  nm <- sub("^upper\\.CL\\.$", "upper.CL", nm)
  names(df) <- nm
  df
}
lsm_ac <- fix_ci_names(lsm_ac)

# Normalize entry labels (handles cases like 'entry12' but is harmless otherwise)
cld_fix <- cld_fix %>%
  mutate(entry = sub("^entry", "", as.character(entry)))

lsm_ac <- lsm_ac %>%
  mutate(entry = as.character(entry))

plot_df <- lsm_ac %>%
  select(entry, estimate, lower.CL, upper.CL) %>%
  inner_join(cld_fix, by = "entry") %>%
  mutate(entry = as.character(entry)) %>%
  arrange(estimate) %>%                      # order by LS-mean
  mutate(entry = factor(entry, levels = entry))

p_cld <- ggplot(plot_df, aes(x = entry, y = estimate)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2) +
  geom_text(aes(label = group, y = upper.CL), vjust = -0.6, size = 3) +
  labs(
    title = "Across-environment LS-means with Tukey CLD",
    x     = "Entry",
    y     = "Adjusted mean yield"
  ) +
  theme_bw()

p_cld

```

Large overlaps in confidence intervals and shared CLD letters indicate similar performance, whereas distinct letters typically reflect statistically distinguishable entries. In practice, you can combine this with an agronomic threshold (for example, ≥ 3 bu/ac) to filter for differences that are both statistically and practically relevant.

---

# 5. Residual spatial diagnostics (single environment)

**Goal.** Inspect whether large‑scale spatial structure remains after the Stage‑1 correction.

```{r heatmap-one-env, fig.height=4.5}
env_example <- trial %>%
  distinct(env) %>%
  slice(1) %>%
  pull(env)

trial_env <- trial %>% filter(env == env_example)

ggplot(trial_env, aes(x = col, y = row, fill = yield)) +
  geom_tile() +
  scale_y_reverse() +
  coord_equal() +
  labs(
    title = paste("Yield heatmap (raw data):", env_example),
    x     = "Column",
    y     = "Row"
  ) +
  theme_bw()
```

Pronounced bands or patches in this heatmap suggest that residual spatial trends remain, even after modeling. The Stage‑1 workflow reduces such structure but may not eliminate it completely, especially if the true process is more complex than the candidate covariance forms.

---

# 6. Variance‑components overview (BLUP model)

**Goal.** Check whether the random‑effect magnitudes in the BLUP model align with simulation choices.

> Note: This chunk re‑fits a mixed model to the simulated trial solely to print the variance components. It does **not** affect the backend output.

```{r varcorr, eval=TRUE}
suppressPackageStartupMessages(library(lme4))

df_blup <- trial %>%
  mutate(
    env   = factor(env),
    rep   = factor(rep),
    entry = factor(entry)
  )

fit2 <- lmer(
  yield ~ 1 + (1 | env) + (1 | env:rep) + (1 | entry) + (1 | entry:env),
  data  = df_blup,
  REML  = TRUE
)

VarCorr(fit2)
```

Very small (near‑zero) variances often produce “boundary (singular) fit” warnings. This is common when the corresponding random effect is weak or the data are limited. Increasing the between‑entry SD, G×E SD, or the number of environments in the simulator will generally stabilize these estimates.

---

# 7. Optional: backend run report

If the backend wrote a `run_report.txt`, it is displayed here for quick reference.

```{r show-run-report, eval=TRUE}
if (!is.null(run_report)) {
  cat(paste(run_report, collapse = "\n"))
} else {
  cat("No run_report.txt found in ../sim_output. Run analyze_trial() with the updated backend to generate one.")
}
```

---

# 8. Summary and guidance

- If rank correlations are high and RMSE is modest, the pipeline is capturing entry signal as intended.  
- If empirical coverage is close to 0.95, the spatial + fallback models’ uncertainty looks reasonably calibrated.  
- If CLD letters align with both statistical and agronomic differences, the reporting is suitable for selection decisions.  
- If spatial residuals appear weak and scattered, Stage‑1 is doing its job; persistent patterns may justify expanding or revising the covariance family set.  
- Singular fits in the BLUP model are warnings, not necessarily failures; they typically mean some random effects are negligible under the current design and simulation parameters.
