---
title: "Variety-Trial Model Validation"
author: "AJ"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: readable
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.align = "center", fig.width = 8, fig.height = 5)
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(readr)
  library(tidyr)
  library(purrr)
  library(stringr)
  library(gridExtra)
})
```

# Purpose

This document validates that the new spatial + multi-environment (MET) workflow estimates what we intend and quantifies how well it performs.

**We check:**
1. Recovery of entry effects and ranks versus known simulation truth.  
2. Stage-1 spatial model selection behavior.  
3. Uncertainty calibration and practical interpretability (CIs, Tukey letters).  
4. Optional diagnostics on variance components and residual structure.

---

# Data Inputs

This validation expects the following files:

- `data/trial_sim.csv`: simulated plots with columns `site, year, env, rep, row, col, entry, yield`.
- `data/entry_truth.csv`: known entry main effects used to simulate data.
- `output/lsm_stage1.csv`: per-environment LS-means and SEs after spatial modeling.
- `output/lsm_across.csv`: inverse-variance weighted LS-means across environments.
- `output/diffs_across.csv`: Tukey-adjusted pairwise comparisons across environments.
- `output/cld_across.csv`: compact letter display (“grouping letters”) across environments.
- `output/entry_blups.csv`: BLUPs of entry effects from the MET model (optional).

```{r load-data}
setwd("C:/Users/ansleybr/Documents/GitHub/sally-jd-yield-analysis")
# Load required files, stop with a clear message if missing
need <- c("data/trial_sim.csv", "data/entry_truth.csv",
          "output/lsm_stage1.csv", "output/lsm_across.csv",
          "output/diffs_across.csv", "output/cld_across.csv",
          "output/entry_blups.csv")

missing <- need[!file.exists(need)]
if (length(missing) > 0) {
  stop(paste("Missing files:\n", paste(missing, collapse = "\n"),
             "\nRun the simulator and analyze_trial() first."))
}

trial   <- readr::read_csv("data/trial_sim.csv", show_col_types = FALSE)
truth   <- readr::read_csv("data/entry_truth.csv", show_col_types = FALSE)
s1      <- readr::read_csv("output/lsm_stage1.csv", show_col_types = FALSE)
lsm_ac  <- readr::read_csv("output/lsm_across.csv", show_col_types = FALSE)
diffs   <- readr::read_csv("output/diffs_across.csv", show_col_types = FALSE)
cld     <- readr::read_csv("output/cld_across.csv", show_col_types = FALSE)
blups   <- readr::read_csv("output/entry_blups.csv", show_col_types = FALSE)

# Normalize entry types
truth <- truth %>% mutate(entry = as.character(entry))
lsm_ac <- lsm_ac %>% mutate(entry = as.character(entry))
blups  <- blups  %>% mutate(entry = as.character(entry))
s1     <- s1     %>% mutate(entry = as.character(entry))
```

---

# 1. Recovery of Entry Effects and Ranks

**Goal,** across environments, entries’ adjusted means and BLUPs should track the true entry effects used in simulation.

**Metrics reported:** Spearman rank correlation (ranking fidelity) and RMSE (numerical accuracy).

```{r recovery-metrics}
cmp_lsm  <- truth %>% inner_join(lsm_ac,  by = "entry")
cmp_blup <- truth %>% inner_join(blups,   by = "entry")

metr <- tibble::tibble(
  source   = c("Across-env LS-means", "BLUPs"),
  rank_cor = c(cor(cmp_lsm$effect, cmp_lsm$estimate, method = "spearman"),
               cor(cmp_blup$effect, cmp_blup$BLUP,   method = "spearman")),
  rmse     = c(sqrt(mean((cmp_lsm$estimate - cmp_lsm$effect)^2)),
               sqrt(mean((cmp_blup$BLUP     - cmp_blup$effect)^2)))
)

metr
```

**Interpretation,** higher rank correlation indicates better retention of order, lower RMSE indicates closer numerical recovery. BLUPs typically stabilize ranks under unbalance.

```{r scatter-plots, fig.height=4.5}
p1 <- ggplot(cmp_lsm, aes(x = effect, y = estimate)) +
  geom_point() + geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Truth vs Across-environment LS-means",
       x = "True entry effect", y = "LS-mean estimate") +
  theme_bw()

p2 <- ggplot(cmp_blup, aes(x = effect, y = BLUP)) +
  geom_point() + geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Truth vs Entry BLUPs",
       x = "True entry effect", y = "BLUP") +
  theme_bw()

gridExtra::grid.arrange(p1, p2, nrow = 1)
```

---

# 2. Stage-1 Spatial Model Selection

**Goal,** AICc chooses plausible covariance families per environment.

```{r stage1-cov-summary}
cov_tab <- s1 %>% count(env, cov) %>% arrange(env, desc(n))
cov_freq <- s1 %>% count(cov) %>% mutate(prop = n / sum(n))
list("Chosen covariance per environment" = cov_tab,
     "Overall frequency" = cov_freq)
```

**Interpretation,** under anisotropic simulation you should often see `expa` selected. If your simulator used isotropic structure, `exp/sph/gau` may be competitive. Consistency across environments indicates stable spatial behavior.

---

# 3. Uncertainty Calibration

**Goal,** per-environment LS-mean 95% CIs should cover the true effect at ≈95%.

```{r coverage}
# Center each environment’s LS-means and compare to entry effects
s1_centered <- s1 |>
  dplyr::group_by(env) |>
  dplyr::mutate(estimate_c = estimate - mean(estimate, na.rm = TRUE)) |>
  dplyr::ungroup()

cov_df <- s1_centered |>
  dplyr::select(env, entry, estimate_c, stderr) |>
  dplyr::inner_join(truth, by = "entry") |>
  dplyr::mutate(
    lo = estimate_c - 1.96 * stderr,
    hi = estimate_c + 1.96 * stderr,
    covered = (effect >= lo) & (effect <= hi)
  )

mean(cov_df$covered, na.rm = TRUE)
```

**Interpretation,** coverage near 0.95 suggests interval calibration is reasonable. Substantially lower coverage may indicate residual underestimation or misspecified spatial form.

```{r}
offset <- mean(lsm_ac$estimate, na.rm = TRUE) - mean(truth$effect, na.rm = TRUE)
with(merge(truth, lsm_ac, by="entry"),
     cor(effect + offset, estimate))  # should be very high
```


---

# 4. Practical Separation: Tukey Letters and Pairwise Differences

**Goal,** the CLD letters summarize Tukey’s multiple comparisons. Entries sharing a letter are **not** significantly different.

```{r cld-merge, fig.height=5}
# Ensure CLD has a 'group' column
if ("group" %in% names(cld)) {
  cld_fix <- cld
} else if (".group" %in% names(cld)) {
  cld_fix <- dplyr::rename(cld, group = .group)
} else {
  stop("CLD table must contain a 'group' or '.group' column.")
}

# Standardize CI column names in case they have trailing dots
fix_ci_names <- function(df){
  nm <- names(df)
  nm <- sub("^lower\\.CL\\.$","lower.CL", nm)
  nm <- sub("^upper\\.CL\\.$","upper.CL", nm)
  names(df) <- nm
  df
}
lsm_ac <- fix_ci_names(lsm_ac)

# Normalize entry labels: "entry12" -> "12"
cld_fix <- cld_fix |>
  dplyr::mutate(entry = sub("^entry", "", as.character(entry)))

# Coerce to character on the other side too
lsm_ac <- lsm_ac |>
  dplyr::mutate(entry = as.character(entry))

plot_df <- lsm_ac |>
  dplyr::select(entry, estimate, lower.CL, upper.CL) |>
  dplyr::inner_join(cld_fix, by = "entry") |>
  dplyr::mutate(entry_num = as.numeric(entry)) |>
  dplyr::arrange(desc(estimate)) |>
  dplyr::mutate(entry = factor(entry, levels = entry))

# If you prefer numeric ordering on the axis, use:
# plot_df <- plot_df |>
#   dplyr::arrange(entry_num) |>
#   dplyr::mutate(entry = factor(entry, levels = unique(entry)))

p <- ggplot2::ggplot(plot_df, ggplot2::aes(x = entry, y = estimate)) +
  ggplot2::geom_point() +
  ggplot2::geom_errorbar(ggplot2::aes(ymin = lower.CL, ymax = upper.CL), width = 0.2) +
  ggplot2::geom_text(ggplot2::aes(label = group, y = upper.CL), vjust = -0.6, size = 3) +
  ggplot2::coord_flip() +
  ggplot2::labs(title = "Across-environment LS-means with Tukey CLD",
                x = "Entry", y = "Adjusted mean") +
  ggplot2::theme_bw()
print(p)
```

**Interpretation,** large overlaps in CIs and shared letters indicate similar performance. You may also apply an agronomic threshold (e.g., ≥ 3 bu/ac) as a practical relevance filter in addition to statistical significance.

---

# 5. Optional: Residual Spatial Diagnostics (Single Environment)

**Goal,** inspect whether large-scale structure remains after spatial correction.

```{r heatmap-one-env, fig.height=4.5}
env_example <- trial %>% distinct(env) %>% slice(1) %>% pull(env)
trial_env <- trial %>% filter(env == env_example)

ggplot(trial_env, aes(x = col, y = row, fill = yield)) +
  geom_tile() + scale_y_reverse() + coord_equal() +
  labs(title = paste("Yield heatmap:", env_example)) +
  theme_bw()
```

**Interpretation,** strong bands or patches would suggest residual spatial trends remain. The Stage-1 model reduces, but might not entirely eliminate, structure—especially if the true process is more complex than the candidate forms.

---

# 6. Optional: Variance Components Overview (BLUP model)

**Goal,** see whether the random-effect magnitudes agree with expectations.

> Note: If your R session hasn’t preserved the `lmer` fit, re-fit briefly to print `VarCorr`.

```{r varcorr, eval=TRUE}
suppressPackageStartupMessages(library(lme4))
df_blup <- trial %>%
  mutate(env = factor(env), rep = factor(rep), entry = factor(entry))
fit2 <- lmer(yield ~ 1 + (1|env) + (1|env:rep) + (1|entry) + (1|entry:env),
             data = df_blup, REML = TRUE)
VarCorr(fit2)
```

**Interpretation,** very small (near-zero) variances produce “boundary (singular) fit” warnings. This is common when the true component is weak or data are limited. Increasing environments, between-entry SD, or G×E SD in the simulator will generally stabilize these estimates.

---

# 7. Summary and Guidance

- **If ranks are recovered** (high Spearman) and **RMSE is modest**, the pipeline is capturing entry signal well.  
- **If coverage is close to 95%**, the spatial models’ uncertainty looks calibrated.  
- **If CLD letters agree with practical differences**, reporting is aligned with agronomic significance.  
- **If spatial residuals look random**, Stage-1 is doing its job; consider expanding the covariance set only if persistent structure remains.  
- **If singular fits occur**, they are warnings, not errors. They indicate some random effects are negligible under current data/parameters; this is acceptable for inference focused on LS-means and Tukey separation, and typically acceptable for BLUP ranking unless pervasive.

---

# Repro Tips

- To test robustness, re-simulate with larger `gxe_sd`, more environments, or higher missingness and rerun this report.  
- Keep raw CSVs and this HTML report together in version control for reproducibility.  
- When moving from simulated to real data, replace `data/entry_truth.csv` checks with cross-validation or split-environment validation metrics (same code structure).
